---
title: "Project_Checkpoint_1"
author: Ian Cay
format: html
---

# Dataset Context

This dataset has political information for 208 countries from 1950-2020. This includes regime information, electoral information, etc.

# Cleaning the Data

```{r}
# Data from the package {democracyData}
# install.packages("pak")
# pak::pak("xmarquez/democracyData")
library(democracyData)
library(tidyverse)

# Raw Dataset
democracy <-
  democracyData::pacl_update

# Cleaned Dataset
democracy_clean <-
  democracyData::pacl_update |> 
  # Changes variable names to snake_case
  janitor::clean_names() |> 
  # Selects and renames variables to include in dataset
  dplyr::select(
    "country_name" = "pacl_update_country",
    "country_code" = "pacl_update_country_isocode",
    "year",
    "regime_category_index" = "dd_regime",
    "regime_category" = "dd_category",
    "is_monarchy" = "monarchy",
    "is_commonwealth" = "commonwealth",
    "monarch_name",
    "monarch_accession_year" = "monarch_accession",
    "monarch_birthyear",
    "is_female_monarch" = "female_monarch",
    "is_democracy" = "democracy",
    "is_presidential" = "presidential",
    "president_name",
    "president_accesion_year" = "president_accesion",
    "president_birthyear",
    "is_interim_phase" = "interim_phase",
    "is_female_president" = "female_president",
    "is_colony" = "colony",
    "colony_of",
    "colony_administrated_by",
    "is_communist" = "communist",
    "has_regime_change_lag" = "regime_change_lag",
    "spatial_democracy",
    "parliament_chambers" = "no_of_chambers_in_parliament",
    "has_proportional_voting" = "proportional_voting",
    "election_system",
    "lower_house_members" = "no_of_members_in_lower_house",
    "upper_house_members" = "no_of_members_in_upper_house",
    "third_house_members" = "no_of_members_in_third_house",
    "has_new_constitution" = "new_constitution",
    "has_full_suffrage" = "fullsuffrage",
    "suffrage_restriction",
    "electoral_category_index" = "electoral",
    "spatial_electoral",
    "has_alternation" = "alternation",
    "is_multiparty" = "multiparty",
    "has_free_and_fair_election" = "free_and_fair_election",
    "parliamentary_election_year",
    "election_month" = "election_month_year",
    "has_postponed_election" = "postponed_election"
  ) |>
  dplyr::mutate(
    election_month = dplyr::na_if(.data$election_month, "?")
  ) |> 
  tidyr::separate_wider_regex(
    "election_month",
    patterns = c(
      election_month = "\\D+",
      election_year = "\\d{4}$"
    ),
    too_few = "align_start"
  ) |> 
  dplyr::mutate(
    electoral_category = dplyr::case_match(
      .data$electoral_category_index,
      0 ~ "no elections",
      1 ~ "single-party elections",
      2 ~ "non-democratic multi-party elections",
      3 ~ "democratic elections"
    ),
    .after = "electoral_category_index"
  ) |> 
  dplyr::mutate(
    #Removes any whitespace in election_month
    election_month = stringr::str_squish(.data$election_month),
    # converts numerical variables to integers
    dplyr::across(
      c(
        tidyselect::ends_with("_index"),
        tidyselect::contains("year"),
        tidyselect::ends_with("_members"),
        "parliament_chambers"
      ),
      as.integer
    ),
    # converts binary variables to logical
    dplyr::across(
      c(
        tidyselect::starts_with("is_"),
        tidyselect::starts_with("has_")
      ),
      as.logical
    )
  )

```

# Research Questions

## With this Dataset

-   Which electoral categories were the most popular over time?
-   How has the popularity of communism changed over time?

## With Supplemental Data

-   Do socialist countries have higher literacy rates over time compared to non-socialist countries?
-   How does the GDP per capita of countries that are democracies change over time compared to countries that aren't democracies?

# Visualization Sketches

```{r}
democracy_clean |>
  # Remove rows where regime_category is missing
  tidyr::drop_na(electoral_category) |>
  # Count the number of countries per regime category per year
  dplyr::count(year, electoral_category) |>
  # Plot the results
  ggplot2::ggplot(
    ggplot2::aes(
      x = year,
      y = n,
      colour = electoral_category
    )
  ) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(
    title = "Popularity of Electoral Categories Over Time",
    x = "Year",
    y = "Number of Countries",
    colour = "Electoral Category"
  ) +
  ggplot2::theme_minimal()
```

```{r}
democracy_clean |>
  # Remove rows where is_communist is missing
  tidyr::drop_na(is_communist) |>
  # Count communist and non-communist countries per year
  dplyr::count(year, is_communist) |>
  # Keep only communist countries
  dplyr::filter(is_communist == TRUE) |>
  # Plot the results
  ggplot2::ggplot(
    ggplot2::aes(
      x = year,
      y = n
    )
  ) +
  ggplot2::geom_line(linewidth = 1, colour = "red") +
  ggplot2::geom_area(alpha = 0.2, fill = "red") +
  ggplot2::labs(
    title = "Number of Communist Countries Over Time",
    x = "Year",
    y = "Number of Communist Countries"
  ) +
  ggplot2::theme_minimal()
```

```{r}
library(ggplot2)
library(png)

visual <- png::readPNG("visual.png")

ggplot2::ggplot() +
  ggplot2::annotation_raster(
    visual,
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = Inf
  )
```
