---
title: "Project 1 Checkpoint 2"
author: "Ian Cay, Drew Cloughley, and Nico Ragasa"
format: html
---

## Data Set Up

```{r}
#| warning: false
library(tidyverse)
library(kableExtra)
library(knitr)

democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

## Research Questions of Interest

-   What regime category is the most prevalent among countries in 2020?
-   Are communist, democratic, or monarchical countries the most prevalent in 2020?
-   Which continent has the highest proportion of countries that have been communist before?

## Number of Countries (Including Colonies) by Regime Category in 2020

```{r}
#| results: 'asis'

count_regime_categories <- function(yr) {
  
  # Filter by the chosen year
  regime_data <- democracy_data |> 
    filter(year == yr, 
           !is.na(regime_category
                  )
           )|> 
    group_by(regime_category) |> 
    summarize(number_of_countries = n(), 
              .groups = "drop"
              ) |> 
  select(regime_category, number_of_countries)

  regime_data |>
    kable() |> 
    kable_styling() |>
    print()
  
  # Create a bar plot
  ggplot(regime_data, 
         aes(x = regime_category, 
             y = number_of_countries, 
             fill = regime_category
             )
         ) +
  geom_col() +
  labs(
    title = paste("Popularity by Regime in", yr),
    x = "Regime Category",
    y = "Number of Countries",
    fill = "Regime Category"
    ) + 
    theme_minimal()
  
}
count_regime_categories(2020)
  
```

## Prevalence of Communism, Monarchies, and Democracies in 2020

```{r}
#| results: 'asis'

count_political_ideologies <- function(selected_year) {
  
  # Filter by the selected year
  filtered_data <- democracy_data |> filter(year == selected_year)
  
  # Iterate over the ideology columns to count TRUE values
  ideology_cols <- c("is_communist", "is_democracy", "is_monarchy")
  ideology_cols
  
  counts <- map_int(ideology_cols, 
                    ~ sum(filtered_data[[.x]], 
                          na.rm = TRUE
                          )
                    )
  
  # Create a summary 
  summary_data <- tibble(
    political_ideology = c("Communism", "Democracy", "Monarchy"),
    number_of_countries = counts,
    year = selected_year
  ) |>
    group_by(political_ideology) |>
    summarize(number_of_countries = number_of_countries) 
  
  summary_data |> 
    kable() |> 
    kable_styling() |> 
    print()
  
  # Create a bar plot
  ggplot(summary_data, 
         aes(x = political_ideology, 
             y = number_of_countries, 
             fill = political_ideology
             )
         ) +
  geom_col() +
  labs(
    title = paste("Popularity of Communism, Democracies, and Monarchies in", selected_year),
    x = "Political Ideology",
    y = "Number of Countries",
    fill = "Political Ideology"
    ) + 
    theme_minimal()

}

count_political_ideologies(2020)

```


## Assigning Continent names to each country and filtering out non-countries

```{r}
#| results: 'asis'

democracy_data <- democracy_data |> 
  mutate(
    continent = case_when(
      
      country_code %in% c(
        "DZA","AGO","BEN","BWA","BFA","BDI","CMR","CPV","CAF","TCD","COM","COG","COD",
        "CIV","DJI","EGY","GNQ","ERI","SWZ","ETH","GAB","GMB","GHA","GIN","GNB","KEN",
        "LSO","LBR","LBY","MDG","MWI","MLI","MRT","MUS","MAR","MOZ","NAM","NER","NGA",
        "RWA","STP","SEN","SYC","SLE","SOM","ZAF","SSD","SDN","TZA","TGO","TUN","UGA",
        "ZMB","ZWE","ZAR"
      ) ~ "Africa",

      # North America (incl. Caribbean)
      country_code %in% c(
        "CAN","USA","MEX","BHS","BRB","CUB","DMA","DOM","GRD","HTI","JAM",
        "KNA","LCA","VCT","TTO","ATG"
      ) ~ "North America",

      # Central America
      country_code %in% c(
        "BLZ","CRI","SLV","GTM","HND","NIC","PAN"
      ) ~ "Central America",

      # South America
      country_code %in% c(
        "ARG","BOL","BRA","CHL","COL","ECU","GUY","PRY","PER","SUR","URY","VEN"
      ) ~ "South America",

      country_code %in% c(
        "AFG","ARM","AZE","BHR","BGD","BTN","BRN","KHM","CHN","CYP","GEO","IND","IDN",
        "IRN","IRQ","ISR","JPN","JOR","KAZ","KWT","KGZ","LAO","LBN","MYS","MDV","MNG",
        "MMR","NPL","PRK","OMN","PAK","PSE","PHL","QAT","SAU","SGP","KOR","LKA","SYR",
        "TWN","TJK","THA","TLS","TUR","TKM","ARE","UZB","VNM","YEM"
      ) ~ "Asia",

      country_code %in% c(
        "ALB","AND","AUT","BLR","BEL","BIH","BGR","HRV","CYP","CZE","DNK","EST","FIN",
        "FRA","DEU","GRC","HUN","ISL","IRL","ITA","LVA","LIE","LTU","LUX","MLT","MDA",
        "MCO","MNE","NLD","MKD","NOR","POL","PRT","ROU","RUS","SMR","SRB","SVK","SVN",
        "ESP","SWE","CHE","UKR","GBR","VAT","GER","ROM"
      ) ~ "Europe",

      country_code %in% c(
        "AUS","FJI","KIR","MHL","FSM","NRU","NZL","PLW","PNG","WSM","SLB","TON","TUV",
        "VUT","NUR"
      ) ~ "Oceania",

      TRUE ~ NA_character_
    )
  ) |> 
  filter(!is.na(continent))
```

## Communist countries by Continent

```{r}
#| results: 'asis'

communist_countries <- democracy_data |> 
  filter(is_communist == TRUE) |> 
  group_by(country_name, continent) |> 
  slice(1) |> 
  ungroup()

communist_summary <- communist_countries |>
  group_by(continent) |> 
  summarize(total_communist_countries = n(), .groups = "drop") |> 
  select(continent, total_communist_countries)

communist_summary |> 
  kable() |>
  kable_styling()

ggplot(communist_summary, aes(x = continent, y = total_communist_countries, fill = continent)) + 
  geom_col() +
  scale_fill_brewer(palette = "Set2") + 
  labs(
    title = "Number of Countries That Have Been Communist by Continent",
    x = "Continent",
    y = "Count"
  )
```

## Percent of Countries that are Communist by Continent

```{r}
#| results: 'asis'

percent_communist <- function(cont){
  democracy_data |>
    filter(continent == cont) |>
    distinct(country_name, is_communist) |>
    summarize(pct_communist = mean(is_communist, na.rm = TRUE)) |>
    pull(pct_communist)
}

percents <- democracy_data |>
  distinct(continent) |>
  pull(continent) |>
  map_dbl(percent_communist)

pct_table <- democracy_data |>
  distinct(continent) |>
  mutate(pct_communist = percents)

pct_table |>
  kable() |>
  kable_styling()

ggplot(pct_table, aes(x = continent, y = pct_communist, fill = continent)) +
  geom_col() +
  labs(x = "Continent", y = "Percent Communist",
       fill = "Continent", title = "Percent of Countries that are Communist by Continent")

```
