---
title: "Project 1 Checkpoint 4"
authors: 
  - name: "Ian Cay"
  - name: "Drew Cloughley"
  - name: "Nico Ragasa"
format: html
embed-resources: true
code-fold: true
theme: flatly
editor: visual
execute: 
  warning: false
---

## Data Set Up

```{r}
#| warning: false
library(tidyverse)
library(ggplot2)
library(ggridges)
library(lmtest)
library(knitr)
library(httr)
library(jsonlite)
library(tidyjson)
library(gt)
library(countrycode)

democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

## Assigning Continent names to each country and filtering out non-countries

```{r}
#| results: 'asis'

democracy_data <- democracy_data |> 
  mutate(
    continent = case_when(
      
      country_code %in% c(
        "DZA","AGO","BEN","BWA","BFA","BDI","CMR","CPV","CAF","TCD","COM","COG","COD",
        "CIV","DJI","EGY","GNQ","ERI","SWZ","ETH","GAB","GMB","GHA","GIN","GNB","KEN",
        "LSO","LBR","LBY","MDG","MWI","MLI","MRT","MUS","MAR","MOZ","NAM","NER","NGA",
        "RWA","STP","SEN","SYC","SLE","SOM","ZAF","SSD","SDN","TZA","TGO","TUN","UGA",
        "ZMB","ZWE","ZAR"
      ) ~ "Africa",

      # North America (incl. Caribbean)
      country_code %in% c(
        "CAN","USA","MEX","BHS","BRB","CUB","DMA","DOM","GRD","HTI","JAM",
        "KNA","LCA","VCT","TTO","ATG"
      ) ~ "North America",

      # Central America
      country_code %in% c(
        "BLZ","CRI","SLV","GTM","HND","NIC","PAN"
      ) ~ "Central America",

      # South America
      country_code %in% c(
        "ARG","BOL","BRA","CHL","COL","ECU","GUY","PRY","PER","SUR","URY","VEN"
      ) ~ "South America",

      country_code %in% c(
        "AFG","ARM","AZE","BHR","BGD","BTN","BRN","KHM","CHN","CYP","GEO","IND","IDN",
        "IRN","IRQ","ISR","JPN","JOR","KAZ","KWT","KGZ","LAO","LBN","MYS","MDV","MNG",
        "MMR","NPL","PRK","OMN","PAK","PSE","PHL","QAT","SAU","SGP","KOR","LKA","SYR",
        "TWN","TJK","THA","TLS","TUR","TKM","ARE","UZB","VNM","YEM"
      ) ~ "Asia",

      country_code %in% c(
        "ALB","AND","AUT","BLR","BEL","BIH","BGR","HRV","CYP","CZE","DNK","EST","FIN",
        "FRA","DEU","GRC","HUN","ISL","IRL","ITA","LVA","LIE","LTU","LUX","MLT","MDA",
        "MCO","MNE","NLD","MKD","NOR","POL","PRT","ROU","RUS","SMR","SRB","SVK","SVN",
        "ESP","SWE","CHE","UKR","GBR","VAT","GER","ROM"
      ) ~ "Europe",

      country_code %in% c(
        "AUS","FJI","KIR","MHL","FSM","NRU","NZL","PLW","PNG","WSM","SLB","TON","TUV",
        "VUT","NUR"
      ) ~ "Oceania",

      TRUE ~ NA_character_
    )
  ) |> 
  filter(!is.na(continent))
```

## Research Questions

-   Are certain political ideologies more popular among higher/lower GDP countries?

-   What is the association between the regime category and population of a country?

-   Do higher/lower population countries prefer certain political ideologies?

## Web Scraping - Population

```{r}
#| label: population-table

library(rvest)

populations <- read_html("https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population")

pop_data <- populations |>
  html_element("table.wikitable") |>
  html_table(fill = TRUE) |>
  select(
    Country = 1, 
    Population = 2
    ) |>
  mutate(
    Country    = trimws(Country),
    Population = as.numeric(gsub("[^0-9]", "", Population))  # strip anything that isn't a digit
  ) |>
  filter(!is.na(Population))
pop_data

```

# 2020 Country Population Data

```{r}
#| label: country-data

pop_data <- pop_data |>
  mutate(
    iso3c = countrycode(Country, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

democracy_data_2020 <- democracy_data |>
  filter(year == 2020) |>
  mutate(
    iso3c = countrycode(country_name, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

# Now join based on iso3c country code
country_data <- pop_data |>
  inner_join(democracy_data_2020, 
             by = "iso3c")
country_data

```

# Ridgeline Plots for the Relationship Between Categorical Variables and Population

```{r}

plot_pop_by_category <- function(cat_var) {
  
  # Store variable names
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "is_communist" = "Communist Status",
                      "is_democracy" = "Democracy Status",
                      "is_monarchy" = "Monarchy Status",
                      .default = var_name)
  
  country_data |>
    filter(!is.na({{ cat_var }})) |> 
    mutate({{ cat_var }} := case_when(
      var_name == "is_communist" & {{ cat_var }} == TRUE  ~ "Communist",
      var_name == "is_communist" & {{ cat_var }} == FALSE ~ "Non-Communist",
      var_name == "is_democracy" & {{ cat_var }} == TRUE  ~ "Democracy",
      var_name == "is_democracy" & {{ cat_var }} == FALSE ~ "Non-Democracy",
      var_name == "is_monarchy" & {{ cat_var }} == TRUE  ~ "Monarchy",
      var_name == "is_monarchy" & {{ cat_var }} == FALSE ~ "Non-Monarchy",
      .default = as.character({{ cat_var }})
    )) |>
    ggplot(aes(y = {{ cat_var }}, x = Population, fill = {{ cat_var }})) + 
    geom_density_ridges(alpha = 0.5) + 
    scale_x_log10(labels = scales::comma) + 
    scale_fill_viridis_d() + 
    labs(
      x     = "Population (log scale)",
      y     = "", 
      title = paste("Country Population by", var_name2), 
      subtitle = paste("Comparing the Populations of Countries by", var_name2)
    ) +
    theme_bw() + 
    theme(legend.position = "none", 
          panel.grid.minor = element_blank(), 
          text = element_text(family = "Helvetica", size = 12))
}

plot_pop_by_category(regime_category)
plot_pop_by_category(is_communist)
plot_pop_by_category(is_democracy)
plot_pop_by_category(is_monarchy)

```

# Summary Statistics for the Relationship Between Categorical Variables and Population

```{r}

summarize_pop_by_category <- function(cat_var) {
  
  # Store variable names for use inside gt()
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "is_communist" = "Communist Status",
                      "is_democracy" = "Democracy Status",
                      "is_monarchy" = "Monarchy Status",
                      .default = var_name)

  country_data |>
    filter(!is.na({{ cat_var }})
    ) |>
    mutate({{ cat_var }} := case_when(
      var_name == "is_communist" & {{ cat_var }} == TRUE  ~ "Communist",
      var_name == "is_communist" & {{ cat_var }} == FALSE ~ "Non-Communist",
      var_name == "is_democracy" & {{ cat_var }} == TRUE  ~ "Democracy",
      var_name == "is_democracy" & {{ cat_var }} == FALSE ~ "Non-Democracy",
      var_name == "is_monarchy" & {{ cat_var }} == TRUE  ~ "Monarchy",
      var_name == "is_monarchy" & {{ cat_var }} == FALSE ~ "Non-Monarchy",
      .default = as.character({{ cat_var }})
    )) |>
    group_by({{ cat_var }}) |>
    summarize(n = n(),
              mean_pop = mean(Population, na.rm = TRUE),
              med_pop = median(Population, na.rm = TRUE),
              sd_pop = sd(Population, na.rm = TRUE),
              se_pop = sd_pop / sqrt(n),
              min_pop = min(Population, na.rm = TRUE),
              max_pop = max(Population, na.rm = TRUE),
              .groups    = "drop"
    ) |>
    arrange(desc(mean_pop)) |>
    gt() |>
    fmt_number(columns = 3:8,
                decimals = 2,
    ) |>
    tab_spanner(label = "Category Breakdown",
                columns = 1:2
    ) |>
    tab_spanner(label = "Midpoints",
                columns = 3:4
    ) |>
    tab_spanner(label = "Variation",
                columns = 5:6
    ) |>
    tab_spanner(label = "Range",
                columns = 7:8
    ) |>
    cols_label(!!var_name := var_name2,
               n = "Count",
               mean_pop = "Mean",
               med_pop = "Median",
               sd_pop = "Standard Deviation",
               se_pop = "Standard Error",
               min_pop = "Minimum",
               max_pop = "Maximum"
    ) |>
    data_color(method = "numeric",
               palette = "Blues"
    ) |>
    tab_header(title = paste("Table of Population Statistics by", var_name2)
    ) |>
    tab_style(style = cell_text(weight = "bold"),
              locations = list(cells_title(groups = "title"), 
                               cells_title(groups = "subtitle"), 
                               cells_column_labels())
    ) |>
    tab_footnote(md("*Note 1*: Standard error was calculated as the standard deviation divided by the square root of the sample size.")) |>
    tab_footnote(md("*Note 2*: Darker blues represent greater values."))
}

summarize_pop_by_category(regime_category)
summarize_pop_by_category(is_communist)
summarize_pop_by_category(is_democracy)
summarize_pop_by_category(is_monarchy)

```

# Statistical Tests for Models Between Categorical Variables and Population

```{r}
# Testing assumptions for model between log(Population) and categorical variables 
pop_assumption_test <- function(cat_var) { 
  
  var_name <- deparse(substitute(cat_var))
  model <- lm(reformulate(var_name, response = "log(Population)"), data = country_data)
  
  par(mfrow = c(2, 2))  # View all 4 diagnostic plots at once 
  plot(model) 
  
  print(shapiro.test(residuals(model)))   # testing normality assumption 
  print(bptest(model))                    # testing equal variance assumption
  
  }

pop_assumption_test(regime_category)
pop_assumption_test(is_communist)
pop_assumption_test(is_democracy)
pop_assumption_test(is_monarchy)

```

```{r}

# Welch's ANOVA
oneway.test(log(Population) ~ regime_category, data = country_data, var.equal = FALSE)

# Mann-Whitney test
wilcox.test(log(Population) ~ is_communist, data = country_data)

# Welch's t-test 
t.test(log(Population) ~ is_democracy, data = country_data, var.equal = FALSE)

# Mann-Whitney test
wilcox.test(log(Population) ~ is_monarchy, data = country_data)

```

# Create Countries Meta Dataset for 2020

```{r}
# Read in csv GDP data
gdp_data <- read.csv("country_gdps.csv")
gdp_data <- gdp_data |>
  select(-X)

# Combine GDP with aggregate 2020 data
country_meta_data <- country_data |>
  inner_join(gdp_data, 
             by = "iso3c", 
             relationship = "many-to-many")

country_meta_data
```

# Ridgeline Plots for the Relationship Between Categorical Variables and GDP in USD (in Billions)

```{r}
plot_gdp_by_category <- function(cat_var) {
  # Store variable names
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "is_communist" = "Communist Status",
                      "is_democracy" = "Democracy Status",
                      "is_monarchy" = "Monarchy Status",
                      .default = var_name)
  
  country_meta_data |>
    filter(!is.na({{ cat_var }})) |> 
    mutate({{ cat_var }} := case_when(
      var_name == "is_communist" & {{ cat_var }} == TRUE  ~ "Communist",
      var_name == "is_communist" & {{ cat_var }} == FALSE ~ "Non-Communist",
      var_name == "is_democracy" & {{ cat_var }} == TRUE  ~ "Democracy",
      var_name == "is_democracy" & {{ cat_var }} == FALSE ~ "Non-Democracy",
      var_name == "is_monarchy" & {{ cat_var }} == TRUE  ~ "Monarchy",
      var_name == "is_monarchy" & {{ cat_var }} == FALSE ~ "Non-Monarchy",
      .default = as.character({{ cat_var }})
    )) |> 
    ggplot(aes(y = {{ cat_var }}, x = GDP, fill = {{ cat_var }})) + 
    geom_density_ridges(alpha = 0.5) + 
    scale_fill_viridis_d() + 
    scale_x_log10(labels = scales::comma) +
    labs(
      x     = "GDP (log scale)",
      y     = "", 
      title = paste("Country GDP in USD (in Billions) by", var_name2), 
      subtitle = paste("Comparing the GDP of Countries by", var_name2)
    ) +
    theme_bw() + 
    theme(legend.position = "none", 
          panel.grid.minor = element_blank(), 
          text = element_text(family = "Helvetica", size = 12))
  
}

plot_gdp_by_category(regime_category)
plot_gdp_by_category(is_communist)
plot_gdp_by_category(is_democracy)
plot_gdp_by_category(is_monarchy)

```

# Summary Statistics for the Relationship Between Categorical Variables and GDP in USD (in Billions)

```{r}

summarize_gdp_by_category <- function(cat_var) {
  
  # Store variable names for use inside gt()
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "is_communist" = "Communist Status",
                      "is_democracy" = "Democracy Status",
                      "is_monarchy" = "Monarchy Status",
                      .default = var_name)

  country_meta_data |>
    filter(!is.na({{ cat_var }})
    ) |>
    mutate({{ cat_var }} := case_when(
      var_name == "is_communist" & {{ cat_var }} == TRUE  ~ "Communist",
      var_name == "is_communist" & {{ cat_var }} == FALSE ~ "Non-Communist",
      var_name == "is_democracy" & {{ cat_var }} == TRUE  ~ "Democracy",
      var_name == "is_democracy" & {{ cat_var }} == FALSE ~ "Non-Democracy",
      var_name == "is_monarchy" & {{ cat_var }} == TRUE  ~ "Monarchy",
      var_name == "is_monarchy" & {{ cat_var }} == FALSE ~ "Non-Monarchy",
      .default = as.character({{ cat_var }})
    )) |>
    group_by({{ cat_var }}) |>
    summarize(n = n(),
              mean_GDP = mean(GDP, na.rm = TRUE),
              med_GDP = median(GDP, na.rm = TRUE),
              sd_GDP = sd(GDP, na.rm = TRUE),
              se_GDP = sd_GDP / sqrt(n),
              min_gdp = min(GDP, na.rm = TRUE),
              max_gdp = max(GDP, na.rm = TRUE),
              .groups    = "drop"
    ) |>
    arrange(desc(mean_GDP)) |>
    gt() |>
    fmt_currency(columns = 3:8,
                 decimals = 2,
                 currency = "USD",
                 placement = "left",
                 pattern = "{x}B"
    ) |>
    tab_spanner(label = "Category Breakdown",
                columns = 1:2
    ) |>
    tab_spanner(label = "Midpoints",
                columns = 3:4
    ) |>
    tab_spanner(label = "Variation",
                columns = 5:6
    ) |>
    tab_spanner(label = "Range",
                columns = 7:8
    ) |>
    cols_label(!!var_name := var_name2,
               n = "Count",
               mean_GDP = "Mean",
               med_GDP = "Median",
               sd_GDP = "Standard Deviation",
               se_GDP = "Standard Error",
               min_gdp = "Minimum",
               max_gdp = "Maximum"
    ) |>
    data_color(method = "numeric",
               palette = "Blues"
    ) |>
    tab_header(title = paste("Table of GDP Statistics by", var_name2),
               subtitle = "(in USD in Billions)"
    ) |>
    tab_style(style = cell_text(weight = "bold"),
              locations = list(cells_title(groups = "title"), 
                               cells_title(groups = "subtitle"), 
                               cells_column_labels())
    ) |>
    tab_footnote(md("*Note 1*: Standard error was calculated as the standard deviation divided by the square root of the sample size.")) |>
    tab_footnote(md("*Note 2*: Darker blues represent greater values."))}

summarize_gdp_by_category(regime_category)
summarize_gdp_by_category(is_communist)
summarize_gdp_by_category(is_democracy)
summarize_gdp_by_category(is_monarchy)

```

# Statistical Tests for Models Between Categorical Variables and GDP

```{r}
# Testing assumptions for model between log(GDP) and categorical variables 
gdp_assumption_test <- function(cat_var) { 
  
  var_name <- deparse(substitute(cat_var))
  model <- lm(reformulate(var_name, response = "log(GDP)"), data = country_meta_data)
  
  par(mfrow = c(2, 2))  # View all 4 diagnostic plots at once 
  plot(model) 
  
  print(shapiro.test(residuals(model)))   # testing normality assumption 
  print(bptest(model))                    # testing equal variance assumption
  
  }

gdp_assumption_test(regime_category)
gdp_assumption_test(is_communist)
gdp_assumption_test(is_democracy)
gdp_assumption_test(is_monarchy)

```

```{r}

# Welch's ANOVA
oneway.test(log(GDP) ~ regime_category, data = country_meta_data, var.equal = FALSE)

# Student's t-test
t.test(log(GDP) ~ is_communist, data = country_meta_data, var.equal = TRUE)

# Welch's t-test 
t.test(log(GDP) ~ is_democracy, data = country_meta_data, var.equal = FALSE)

# Student's t-test
t.test(log(GDP) ~ is_monarchy, data = country_meta_data, var.equal = TRUE)

```
