---
title: "Project 1 Checkpoint 3"
authors: 
  - name: "Ian Cay"
  - name: "Drew Cloughley"
  - name: "Nico Ragasa"
format: html
editor: visual
execute: 
  warning: false
---

## Data Set Up

```{r}
#| warning: false
library(tidyverse)
library(kableExtra)
library(knitr)
library(httr)
library(jsonlite)
library(tidyjson)
library(glue)
library(gt)
library(countrycode)

democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

## Assigning Continent names to each country and filtering out non-countries

```{r}
#| results: 'asis'

democracy_data <- democracy_data |> 
  mutate(
    continent = case_when(
      
      country_code %in% c(
        "DZA","AGO","BEN","BWA","BFA","BDI","CMR","CPV","CAF","TCD","COM","COG","COD",
        "CIV","DJI","EGY","GNQ","ERI","SWZ","ETH","GAB","GMB","GHA","GIN","GNB","KEN",
        "LSO","LBR","LBY","MDG","MWI","MLI","MRT","MUS","MAR","MOZ","NAM","NER","NGA",
        "RWA","STP","SEN","SYC","SLE","SOM","ZAF","SSD","SDN","TZA","TGO","TUN","UGA",
        "ZMB","ZWE","ZAR"
      ) ~ "Africa",

      # North America (incl. Caribbean)
      country_code %in% c(
        "CAN","USA","MEX","BHS","BRB","CUB","DMA","DOM","GRD","HTI","JAM",
        "KNA","LCA","VCT","TTO","ATG"
      ) ~ "North America",

      # Central America
      country_code %in% c(
        "BLZ","CRI","SLV","GTM","HND","NIC","PAN"
      ) ~ "Central America",

      # South America
      country_code %in% c(
        "ARG","BOL","BRA","CHL","COL","ECU","GUY","PRY","PER","SUR","URY","VEN"
      ) ~ "South America",

      country_code %in% c(
        "AFG","ARM","AZE","BHR","BGD","BTN","BRN","KHM","CHN","CYP","GEO","IND","IDN",
        "IRN","IRQ","ISR","JPN","JOR","KAZ","KWT","KGZ","LAO","LBN","MYS","MDV","MNG",
        "MMR","NPL","PRK","OMN","PAK","PSE","PHL","QAT","SAU","SGP","KOR","LKA","SYR",
        "TWN","TJK","THA","TLS","TUR","TKM","ARE","UZB","VNM","YEM"
      ) ~ "Asia",

      country_code %in% c(
        "ALB","AND","AUT","BLR","BEL","BIH","BGR","HRV","CYP","CZE","DNK","EST","FIN",
        "FRA","DEU","GRC","HUN","ISL","IRL","ITA","LVA","LIE","LTU","LUX","MLT","MDA",
        "MCO","MNE","NLD","MKD","NOR","POL","PRT","ROU","RUS","SMR","SRB","SVK","SVN",
        "ESP","SWE","CHE","UKR","GBR","VAT","GER","ROM"
      ) ~ "Europe",

      country_code %in% c(
        "AUS","FJI","KIR","MHL","FSM","NRU","NZL","PLW","PNG","WSM","SLB","TON","TUV",
        "VUT","NUR"
      ) ~ "Oceania",

      TRUE ~ NA_character_
    )
  ) |> 
  filter(!is.na(continent))
```

## Research Questions

-   Are certain political ideologies more popular among higher/lower GDP countries?

-   What is the association between the regime category and population of a country?

-   Do higher/lower population countries prefer certain political ideologies?

## To-Do

two new country-level data sources

Stat 431: One of these must be acquired using APIs or webscraping

Stat 541: One of these must be acquired using APIs, the other using webscraping.

descriptions of each of the new datasets

a “meta” dataset containing all the acquired data (joined together)

at least two data summaries OR visualizations incorporating the additional country-level data

## Web Scraping - Population

```{r}
#| label: population-table

library(rvest)

populations <- read_html("https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population")

pop_data <- populations |>
  html_element("table.wikitable") |>
  html_table(fill = TRUE) |>
  select(
    Country = 1, 
    Population = 2
    ) |>
  mutate(
    Country    = trimws(Country),
    Population = as.numeric(gsub("[^0-9]", "", Population))  # strip anything that isn't a digit
  ) |>
  filter(!is.na(Population))
pop_data

```

# 2020 Country Population Data

```{r}
#| label: country-data

pop_data <- pop_data |>
  mutate(
    iso3c = countrycode(Country, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

democracy_data_2020 <- democracy_data |>
  filter(year == 2020) |>
  mutate(
    iso3c = countrycode(country_name, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

# Now join based on iso3c country code
country_data <- pop_data |>
  inner_join(democracy_data_2020, 
             by = "iso3c")
country_data

```

# Box-plots for the Relationship Between Categorical Variables and Population

```{r}

plot_pop_by_category <- function(cat_var) {
  ggplot(country_data, aes(y = {{ cat_var }}, x = Population, fill = {{ cat_var }})) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 1.5, alpha = 0.7) +
    scale_x_log10(labels = scales::comma) +
    labs(
      title = paste("Country Population by", deparse(substitute(cat_var))),
      y     = deparse(substitute(cat_var)),
      x     = "Population (log scale)"
    ) +
    theme_minimal()
}

plot_pop_by_category(regime_category)
plot_pop_by_category(is_communist)
plot_pop_by_category(is_democracy)
plot_pop_by_category(is_monarchy)

```

# Summary Statistics for the Relationship Between Categorical Variables and Population


```{r}

summarize_pop_by_category <- function(cat_var) {
  country_data |>
    group_by({{ cat_var }}) |>
    summarize(
      n          = n(),
      mean_pop   = mean(Population, na.rm = TRUE),
      median_pop = median(Population, na.rm = TRUE),
      sd_pop     = sd(Population, na.rm = TRUE),
      min_pop    = min(Population, na.rm = TRUE),
      max_pop    = max(Population, na.rm = TRUE),
      .groups    = "drop"
    ) |>
    arrange(desc(median_pop))
}

summarize_pop_by_category(regime_category)
summarize_pop_by_category(is_communist)
summarize_pop_by_category(is_democracy)
summarize_pop_by_category(is_monarchy)

```

## API - GDP in USD

```{r}
#| label: country-api

# Obtain 2020 GDP in USD for inputted country
get_gdp_2020 <- function(country, key = my_key)
  {
  
  # Store url to access API with country and key queries
  url <- glue("https://api.api-ninjas.com/v1/gdp?year=2020;country={country};X-Api-Key={key}")
  
  # Temporarily store JSON data safely
  my_data <- safely(fromJSON)(url)
  
  # Check if JSON was successfully obtained
  if (is.null(my_data$result) |
      length(my_data$result) == 0) {
    return(NA)
  } else {
    return(my_data$result$gdp_nominal)
  }
}

```

```{r}
#| echo: false
my_key <- "Zk9p4D2UR8cKVCa1GKMhJgqnWRTzotRYLJOlagkr"
```

# Store Country GDPs in USD

```{r}
#| eval: false
# We used a personal API key stored as "my_key"

# Get GDP for all countries
gdps <- map_dbl(.x = country_data$iso3c, .f = get_gdp)

# Store GDP with country code as a tibble
country_gdps <- tibble(iso3c = country_data$iso3c, GDP = gdps)

# Save tibble as a csv
write.csv(country_gdps, file = "country_gdps.csv")
```

# Create Countries Meta Dataset for 2020

```{r}
# Read in csv GDP data
gdp_data <- read.csv("country_gdps.csv")
gdp_data <- gdp_data |>
  select(-X)

# Combine GDP with aggregate 2020 data
country_meta_data <- country_data |>
  inner_join(gdp_data, 
             by = "iso3c")

country_meta_data
```

# Box-plots for the Relationship Between Categorical Variables and GDP

```{r}
plot_gdp_by_category <- function(cat_var) {
  ggplot(country_meta_data, aes(y = {{ cat_var }}, x = GDP, fill = {{ cat_var }})) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 1.5, alpha = 0.7) +
    scale_x_log10(labels = scales::comma) +
    labs(
      title = paste("Country GDP in USD by", deparse(substitute(cat_var))),
      y     = deparse(substitute(cat_var)),
      x     = "GDP (log scale)"
    ) +
    theme_minimal()
}

plot_gdp_by_category(regime_category)
plot_gdp_by_category(is_communist)
plot_gdp_by_category(is_democracy)
plot_gdp_by_category(is_monarchy)

```

# Summary Statistics for the Relationship Between Categorical Variables and GDP

```{r}
summarize_gdp_by_category <- function(cat_var) {
  country_meta_data |>
    group_by({{ cat_var }}) |>
    summarize(
      n          = n(),
      mean_gdp   = mean(GDP, na.rm = TRUE),
      median_gdp = median(GDP, na.rm = TRUE),
      sd_gdp     = sd(GDP, na.rm = TRUE),
      min_gdp    = min(GDP, na.rm = TRUE),
      max_gdp    = max(GDP, na.rm = TRUE),
      .groups    = "drop"
    ) |>
    arrange(desc(median_gdp))
}

summarize_gdp_by_category(regime_category)
summarize_gdp_by_category(is_communist)
summarize_gdp_by_category(is_democracy)
summarize_gdp_by_category(is_monarchy)

```
