---
title: "Project 1 Checkpoint 4"
authors: 
  - name: "Ian Cay"
  - name: "Drew Cloughley"
  - name: "Nico Ragasa"
format: html
embed-resources: true
code-fold: true
theme: flatly
editor: visual
execute: 
  warning: false
---

## Data Set Up

```{r}
#| warning: false
library(tidyverse)
library(ggplot2)
library(ggridges)
library(kableExtra)
library(knitr)
library(httr)
library(jsonlite)
library(tidyjson)
library(countrycode)

democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

## Assigning Continent names to each country and filtering out non-countries

```{r}
#| results: 'asis'

democracy_data <- democracy_data |> 
  mutate(
    continent = case_when(
      
      country_code %in% c(
        "DZA","AGO","BEN","BWA","BFA","BDI","CMR","CPV","CAF","TCD","COM","COG","COD",
        "CIV","DJI","EGY","GNQ","ERI","SWZ","ETH","GAB","GMB","GHA","GIN","GNB","KEN",
        "LSO","LBR","LBY","MDG","MWI","MLI","MRT","MUS","MAR","MOZ","NAM","NER","NGA",
        "RWA","STP","SEN","SYC","SLE","SOM","ZAF","SSD","SDN","TZA","TGO","TUN","UGA",
        "ZMB","ZWE","ZAR"
      ) ~ "Africa",

      # North America (incl. Caribbean)
      country_code %in% c(
        "CAN","USA","MEX","BHS","BRB","CUB","DMA","DOM","GRD","HTI","JAM",
        "KNA","LCA","VCT","TTO","ATG"
      ) ~ "North America",

      # Central America
      country_code %in% c(
        "BLZ","CRI","SLV","GTM","HND","NIC","PAN"
      ) ~ "Central America",

      # South America
      country_code %in% c(
        "ARG","BOL","BRA","CHL","COL","ECU","GUY","PRY","PER","SUR","URY","VEN"
      ) ~ "South America",

      country_code %in% c(
        "AFG","ARM","AZE","BHR","BGD","BTN","BRN","KHM","CHN","CYP","GEO","IND","IDN",
        "IRN","IRQ","ISR","JPN","JOR","KAZ","KWT","KGZ","LAO","LBN","MYS","MDV","MNG",
        "MMR","NPL","PRK","OMN","PAK","PSE","PHL","QAT","SAU","SGP","KOR","LKA","SYR",
        "TWN","TJK","THA","TLS","TUR","TKM","ARE","UZB","VNM","YEM"
      ) ~ "Asia",

      country_code %in% c(
        "ALB","AND","AUT","BLR","BEL","BIH","BGR","HRV","CYP","CZE","DNK","EST","FIN",
        "FRA","DEU","GRC","HUN","ISL","IRL","ITA","LVA","LIE","LTU","LUX","MLT","MDA",
        "MCO","MNE","NLD","MKD","NOR","POL","PRT","ROU","RUS","SMR","SRB","SVK","SVN",
        "ESP","SWE","CHE","UKR","GBR","VAT","GER","ROM"
      ) ~ "Europe",

      country_code %in% c(
        "AUS","FJI","KIR","MHL","FSM","NRU","NZL","PLW","PNG","WSM","SLB","TON","TUV",
        "VUT","NUR"
      ) ~ "Oceania",

      TRUE ~ NA_character_
    )
  ) |> 
  filter(!is.na(continent))
```

## Research Questions

-   Are certain political ideologies more popular among higher/lower GDP countries?

-   What is the association between the regime category and population of a country?

-   Do higher/lower population countries prefer certain political ideologies?

## To-Do

two professionally formatted visualizations every visualization must have alternative (alt) text, non-default colors, a non-default theme, and clear labels at least one visualization must use a more complex geometry (e.g., geom_segment(), geom_density_ridge() geom_ribbon(), geom_tile(), geom_hex()) fits one statistical model addressing one of the research questions posed professionally formats the output of that model to communicate the results Formatted values (e.g., decimals, percentages, currencies) Boldface headers Captions

## Web Scraping - Population

```{r}
#| label: population-table

library(rvest)

populations <- read_html("https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population")

pop_data <- populations |>
  html_element("table.wikitable") |>
  html_table(fill = TRUE) |>
  select(
    Country = 1, 
    Population = 2
    ) |>
  mutate(
    Country    = trimws(Country),
    Population = as.numeric(gsub("[^0-9]", "", Population))  # strip anything that isn't a digit
  ) |>
  filter(!is.na(Population))
pop_data

```

# 2020 Country Population Data

```{r}
#| label: country-data

pop_data <- pop_data |>
  mutate(
    iso3c = countrycode(Country, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

democracy_data_2020 <- democracy_data |>
  filter(year == 2020) |>
  mutate(
    iso3c = countrycode(country_name, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

# Now join based on iso3c country code
country_data <- pop_data |>
  inner_join(democracy_data_2020, 
             by = "iso3c")
country_data

```

# Ridgeline Plot for the Relationship between Regime Category and Population

```{r}

country_data |>
  filter(!is.na(is_monarchy)) |> 
  ggplot(aes(x = Population, 
             y = is_monarchy, 
             fill = is_monarchy)) + 
  geom_density_ridges(alpha = 0.5) +
  scale_x_log10(labels = scales::comma) +
  theme_light() + 
  theme(legend.position = "none",
        text = element_text(family = "Times", size = 12)) +
  scale_fill_brewer(palette = "Accent") +
  labs(
    title = "Population Ridgeline Densities Between Monarchies and Non-monarchies",
    x = "Population (log scale)",
    y = ""
  )
  
```

# Two-Sample t-test for Population and Communism

```{r}

t.test(Population ~ is_monarchy, data = country_meta_data)

```

# Summary Statistics for the Relationship Between Categorical Variables and Population

```{r}

summarize_pop_by_category <- function(cat_var) {
  country_data |>
    group_by({{ cat_var }}) |>
    summarize(
      n          = n(),
      mean_pop   = mean(Population, na.rm = TRUE),
      median_pop = median(Population, na.rm = TRUE),
      sd_pop     = sd(Population, na.rm = TRUE),
      min_pop    = min(Population, na.rm = TRUE),
      max_pop    = max(Population, na.rm = TRUE),
      .groups    = "drop"
    ) |>
    arrange(desc(median_pop))
}

summarize_pop_by_category(regime_category)
summarize_pop_by_category(is_communist)
summarize_pop_by_category(is_democracy)
summarize_pop_by_category(is_monarchy)

```

## API - GDP in USD

```{r}
#| label: country-api

# Obtain 2020 GDP in USD for inputted country
get_gdp_2020 <- function(country, key = my_key)
  {
  
  # Store url to access API with country and key queries
  url <- glue("https://api.api-ninjas.com/v1/gdp?year=2020;country={country};X-Api-Key={key}")
  
  # Temporarily store JSON data safely
  my_data <- safely(fromJSON)(url)
  
  # Check if JSON was successfully obtained
  if (is.null(my_data$result) |
      length(my_data$result) == 0) {
    return(NA)
  } else {
    return(my_data$result$gdp_nominal)
  }
}

```


# Create Countries Meta Dataset for 2020

```{r}
# Read in csv GDP data
gdp_data <- read.csv("country_gdps.csv")
gdp_data <- gdp_data |>
  select(-X)

# Combine GDP with aggregate 2020 data
country_meta_data <- country_data |>
  inner_join(gdp_data, 
             by = "iso3c")

country_meta_data
```

# Box-plots for the Relationship Between Categorical Variables and GDP in USD (in Billions)

```{r}
plot_gdp_by_category <- function(cat_var) {
  ggplot(country_meta_data, aes(y = {{ cat_var }}, x = GDP, fill = {{ cat_var }})) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 1.5, alpha = 0.7) +
    scale_x_log10(labels = scales::comma) +
    labs(
      title = paste("Country GDP in USD (in Billions) by", deparse(substitute(cat_var))),
      y     = deparse(substitute(cat_var)),
      x     = "GDP (log scale)"
    ) +
    theme_minimal()
}

plot_gdp_by_category(regime_category)
plot_gdp_by_category(is_communist)
plot_gdp_by_category(is_democracy)
plot_gdp_by_category(is_monarchy)

```

# Summary Statistics for the Relationship Between Categorical Variables and GDP in USD (in Billions)

```{r}
summarize_gdp_by_category <- function(cat_var) {
  country_meta_data |>
    group_by({{ cat_var }}) |>
    summarize(
      n          = n(),
      mean_gdp   = mean(GDP, na.rm = TRUE),
      median_gdp = median(GDP, na.rm = TRUE),
      sd_gdp     = sd(GDP, na.rm = TRUE),
      min_gdp    = min(GDP, na.rm = TRUE),
      max_gdp    = max(GDP, na.rm = TRUE),
      .groups    = "drop"
    ) |>
    arrange(desc(median_gdp))
}

summarize_gdp_by_category(regime_category)
summarize_gdp_by_category(is_communist)
summarize_gdp_by_category(is_democracy)
summarize_gdp_by_category(is_monarchy)

```
