---
title: "Analyzing Factors of Different Political Structure"
subtitle: "What effect does political ideology have on population and nominal GDP?"
authors: 
  - name: "Ian Cay"
  - name: "Drew Cloughley"
  - name: "Nico Ragasa"
format: html
embed-resources: true
code-fold: true
theme: flatly
editor: visual
execute: 
  warning: false
---

China stands apart from most of the world's nations in a striking way: while liberal democracies dominate the global political landscape, China remains as one of the few communist countries, and despite its differences, it ranks near the top of in both nominal GDP and population.

That scale, however, comes with immense pressure on resources, which is part of what drove China to introduce the One-Child Policy from 1980 to 2015, which limited most families to a single child. While the policy eased strain on resources in the short term, it gradually hollowed out China's working-age population, creating economic challenges that persist today.

This raises a broader question worth examining: do a country's political structures influence its population size? And if so, what does that mean for economic growth?

## The Data Collection Process

To investigate this, we first utilized the dataset titled *Democracy and Dictatorship*, which consists of information on the political structures of countries from 1950-2020. This dataset was collected from databases created by Christian Bjørnskov and Martin Rode in 2020, and it was published by Jon Harmon for *tidytuesday* on GitHub on November 5, 2024.

Afterwards, to collect the populations of each country, we web scraped from Wikipedia page *List of countries and dependencies by population* (https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population), which were obtained from the United Nations and the United Nations Department of Economic and Social Affairs. The date in which this population data was collected varies by country, but it was collected from 2018 to 2026. Due to 2020 population statistics not being available for all countries, we obtained the closest value to reflect the 2020-era population.

Finally, we obtained the GDP of each country through an API request to *API Ninjas*, sourced from the *World Bank*. The data contains GDP statistics dating back to 1980, however the publication date of the API dataset is not listed. We retrieved values of country GDP through a function with an API request for an inputted country with the query parameter of *year* set to 2020. The function outputted the country's nominal GDP in billions in US dollars. Each request used a personal, hidden key, and the function was iterated over every country in our data, stored in a temporary R data frame, and appended to our meta dataset for 2020 using each countries ISO alpha-3 code.

```{r}
#| warning: false

library(tidyverse)
library(ggplot2)
library(ggridges)
library(lmtest)
library(gt)

democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

```{r}
#| label: population-table

library(rvest)

populations <- read_html("https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population")

pop_data <- populations |>
  html_element("table.wikitable") |>
  html_table(fill = TRUE) |>
  select(
    Country = 1, 
    Population = 2
    ) |>
  mutate(
    Country    = trimws(Country),
    Population = as.numeric(gsub("[^0-9]", "", Population))  # strip anything that isn't a digit
  ) |>
  filter(!is.na(Population))

```

```{r}
#| label: country-data
#| warning: false

library(countrycode)

pop_data <- pop_data |>
  mutate(
    iso3c = countrycode(Country, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

democracy_data_2020 <- democracy_data |>
  filter(year == 2020) |>
  mutate(
    iso3c = countrycode(country_name, 
                        origin = "country.name", 
                        destination = "iso3c"
                        ))

# Now join based on iso3c country code
country_data <- pop_data |>
  inner_join(democracy_data_2020, 
             by = "iso3c")

```

```{r}
#| label: country-api

library(glue)
library(jsonlite)

# Obtain 2020 GDP in USD for inputted country
get_gdp_2020 <- function(country, key = my_key)
  {
  
  # Store url to access API with country and key queries
  url <- glue("https://api.api-ninjas.com/v1/gdp?year=2020;country={country};X-Api-Key={key}")
  
  # Temporarily store JSON data safely
  my_data <- safely(fromJSON)(url)
  
  # Check if JSON was successfully obtained
  if (is.null(my_data$result) |
      length(my_data$result) == 0) {
    return(NA)
  } else {
    return(my_data$result$gdp_nominal)
  }
}

```

```{r}
#| echo: false
my_key <- "Zk9p4D2UR8cKVCa1GKMhJgqnWRTzotRYLJOlagkr"

```

```{r}
#| eval: false
# We used a personal API key stored as "my_key"

# Get GDP for all countries
gdps <- map_dbl(.x = country_data$iso3c, .f = get_gdp_2020)

# Store GDP with country code as a tibble
country_gdps <- tibble(iso3c = country_data$iso3c, GDP = gdps)

# Save tibble as a csv
write.csv(country_gdps, file = "country_gdps.csv")

```

```{r}

# Read in csv GDP data
gdp_data <- read.csv("country_gdps.csv")
gdp_data <- gdp_data |>
  select(-X)

# Combine GDP with aggregate 2020 data
country_meta_data <- country_data |>
  inner_join(gdp_data, 
             by = "iso3c", 
             relationship = "many-to-many")

```

```{r}
country_meta_data <- country_meta_data |>
  filter(!is.na(regime_category_index)) |> 
  mutate(gov_type = case_when(
    is_communist == TRUE ~ "Communist",
    is_democracy == TRUE & is_monarchy == TRUE ~ "Democratic Monarchy",
    is_democracy == TRUE & is_monarchy == FALSE ~ "Pure Democracy",
    is_democracy == FALSE & is_monarchy == TRUE ~ "Pure Monarchy",
    TRUE ~ "Other"), 
    regime_category = case_when(
    regime_category_index == 0 ~ "Parliamentary democracy",
    regime_category_index == 1 ~ "Mixed democratic", 
    regime_category_index == 2 ~ "Presidential democracy",
    regime_category_index == 3 ~ "Civilian dictatorship", 
    regime_category_index == 4 ~ "Military dictatorship",
    regime_category_index == 5 ~ "Royal dictatorship")
  )
```

```{r}

plot_gdp_by_category <- function(cat_var) {
  # Store variable names
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "gov_type" = "Government Type",
                      .default = var_name)
  
  country_meta_data |>
    filter(!is.na({{ cat_var }})) |> 
    ggplot(aes(y = {{ cat_var }}, x = GDP, fill = {{ cat_var }})) + 
    geom_density_ridges(alpha = 0.5) + 
    scale_fill_viridis_d() + 
    scale_x_log10(labels = scales::comma) +
    labs(
      x     = "GDP (log scale)",
      y     = "", 
      title = paste("Country GDP in USD (in Billions) by", var_name2), 
      subtitle = paste("Comparing the GDP of Countries by", var_name2)
    ) +
    theme_bw() + 
    theme(legend.position = "none", 
          panel.grid.minor = element_blank(), 
          text = element_text(family = "Helvetica", size = 12))
  
}

```

```{r}

summarize_gdp_by_category <- function(cat_var) {
  
  # Store variable names for use inside gt()
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "gov_type" = "Government Type",
                      .default = var_name)

  country_meta_data |>
    filter(!is.na({{ cat_var }}), 
           !is.na(GDP)
    ) |>
    group_by({{ cat_var }}) |>
    summarize(n = n(),
              mean_GDP = mean(GDP, na.rm = TRUE),
              med_GDP = median(GDP, na.rm = TRUE),
              sd_GDP = sd(GDP, na.rm = TRUE),
              se_GDP = sd_GDP / sqrt(n),
              min_gdp = min(GDP, na.rm = TRUE),
              max_gdp = max(GDP, na.rm = TRUE),
              .groups    = "drop"
    ) |>
    arrange(desc(mean_GDP)) |>
    gt() |>
    fmt_currency(columns = 3:8,
                 decimals = 2,
                 currency = "USD",
                 placement = "left",
                 pattern = "{x}B"
    ) |>
    tab_spanner(label = "Category Breakdown",
                columns = 1:2
    ) |>
    tab_spanner(label = "Midpoints",
                columns = 3:4
    ) |>
    tab_spanner(label = "Variation",
                columns = 5:6
    ) |>
    tab_spanner(label = "Range",
                columns = 7:8
    ) |>
    cols_label(!!var_name := var_name2,
               n = "Count",
               mean_GDP = "Mean",
               med_GDP = "Median",
               sd_GDP = "Standard Deviation",
               se_GDP = "Standard Error",
               min_gdp = "Minimum",
               max_gdp = "Maximum"
    ) |>
    data_color(columns = 2:8, 
               method = "numeric",
               palette = "Blues"
    ) |>
    tab_header(title = paste("Table of GDP Statistics by", var_name2),
               subtitle = "(in USD in Billions)"
    ) |>
    tab_style(style = cell_text(weight = "bold"),
              locations = list(cells_title(groups = "title"), 
                               cells_title(groups = "subtitle"), 
                               cells_column_labels())
    ) |>
    tab_footnote(md("*Note 1*: Standard error was calculated as the standard deviation divided by the square root of the sample size.")) |>
    tab_footnote(md("*Note 2*: Darker blues represent greater values."))
  }

```

```{r}

plot_pop_by_category <- function(cat_var) {
  
  # Store variable names
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "gov_type" = "Government Type",
                      .default = var_name)
  
  country_meta_data |>
    filter(!is.na({{ cat_var }}), !is.na(Population)) |> 
    ggplot(aes(y = {{ cat_var }}, x = Population, fill = {{ cat_var }})) + 
    geom_density_ridges(alpha = 0.5) + 
    scale_x_log10(labels = scales::comma) + 
    scale_fill_viridis_d() + 
    labs(
      x     = "Population (log scale)",
      y     = "", 
      title = paste("Country Population by", var_name2), 
      subtitle = paste("Comparing the Populations of Countries by", var_name2)
    ) +
    theme_bw() + 
    theme(legend.position = "none", 
          panel.grid.minor = element_blank(), 
          text = element_text(family = "Helvetica", size = 12))
}
```

```{r}

summarize_pop_by_category <- function(cat_var) {
  
  # Store variable names for use inside gt()
  var_name <- deparse(substitute(cat_var))
  var_name2 <- recode(var_name,
                      "regime_category" = "Regime Category",
                      "gov_type" = "Government Type",
                      .default = var_name)

  country_meta_data |>
    filter(!is.na({{ cat_var }}), 
           !is.na(Population)
    ) |>
    group_by({{ cat_var }}) |>
    summarize(n = n(),
              mean_pop = mean(Population, na.rm = TRUE),
              med_pop = median(Population, na.rm = TRUE),
              sd_pop = sd(Population, na.rm = TRUE),
              se_pop = sd_pop / sqrt(n),
              min_pop = min(Population, na.rm = TRUE),
              max_pop = max(Population, na.rm = TRUE),
              .groups    = "drop"
    ) |>
    arrange(desc(mean_pop)) |>
    gt() |>
    fmt_number(columns = 3:8,
                decimals = 2
    ) |>
    tab_spanner(label = "Category Breakdown",
                columns = 1:2
    ) |>
    tab_spanner(label = "Midpoints",
                columns = 3:4
    ) |>
    tab_spanner(label = "Variation",
                columns = 5:6
    ) |>
    tab_spanner(label = "Range",
                columns = 7:8
    ) |>
    cols_label(!!var_name := var_name2,
               n = "Count",
               mean_pop = "Mean",
               med_pop = "Median",
               sd_pop = "Standard Deviation",
               se_pop = "Standard Error",
               min_pop = "Minimum",
               max_pop = "Maximum"
    ) |>
    data_color(columns = 2:8, 
               method = "numeric",
               palette = "Blues"
    ) |>
    tab_header(title = paste("Table of Population Statistics by", var_name2)
    ) |>
    tab_style(style = cell_text(weight = "bold"),
              locations = list(cells_title(groups = "title"), 
                               cells_title(groups = "subtitle"), 
                               cells_column_labels())
    ) |>
    tab_footnote(md("*Note 1*: Standard error was calculated as the standard deviation divided by the square root of the sample size.")) |>
    tab_footnote(md("*Note 2*: Darker blues represent greater values."))
}
```

## An Analysis of Populations Across Political Ideologies

```{r}
#| fig-alt: "Ridgeline plots showing distributions of log(Population) for different regime categories. 'Communist' appears to have the highest median population, and 'Democratic monarchy' seems to have the lowest median population."
plot_pop_by_category(gov_type)
```

The distributions of population for pure monarchies, pure democracies, and other government types appear to be normal, while democratic monarchies and communist governments have a bimodal distribution. Communist governments seem to have the largest average population, given the right-most center. Communist countries show a staggeringly greater average population than other government types, at over 220 million, and the lowest average population is democratic monarchies. The standard error for communist governments is large, so we need to perform a test of significance to determine if the difference is statistically significant.

```{r}
summarize_pop_by_category(regime_category)
```

The distributions of populations across regime categories appear to be approximately normal, with the exception of parliamentary democracies. Civilian dictatorships or presidential democracies appear to have the largest populations on average, and royal dictatorships appear to have the least. Civilian dictatorships have the largest mean population, followed by presidential democracies and parliamentary democracies. Royal dictatorships have the lowest average population.

```{r}
# Welch's ANOVA between regime_category and log(Population)
pop_regime_model <- oneway.test(log(Population) ~ regime_category, data = country_meta_data, var.equal = FALSE)

data.frame(
    f_stat   = unname(pop_regime_model$statistic),
    num_df   = unname(pop_regime_model$parameter[1]),
    denom_df = unname(pop_regime_model$parameter[2]),
    p_value  = pop_regime_model$p.value
    ) |>
  gt() |> 
  cols_label(
    f_stat   = "F-statistic", 
    num_df   = "Numerator Degrees of Freedom", 
    denom_df = "Denominator Degrees of Freedom", 
    p_value  = "P-value"
    ) |> 
  tab_header(title = paste("Welch's ANOVA for log(Population) and Regime Category")) |> 
  tab_style(style = cell_text(weight = "bold"), 
            locations = list(cells_title(groups = "title"), 
                             cells_column_labels()
                             )
            )
```

At an ⍺ = 0.05 significance level, we have extremely strong evidence that there is an association between the geometric mean population and the regime categories of countries.

```{r}
# Welch's ANOVA between government type and log(Population)
pop_gov_model <- oneway.test(log(Population) ~ gov_type, data = country_meta_data, var.equal = FALSE)

data.frame(
    f_stat   = unname(pop_gov_model$statistic),
    num_df   = unname(pop_gov_model$parameter[1]),
    denom_df = unname(pop_gov_model$parameter[2]),
    p_value  = pop_gov_model$p.value
    ) |>
  gt() |> 
  cols_label(
    f_stat   = "F-statistic", 
    num_df   = "Numerator Degrees of Freedom", 
    denom_df = "Denominator Degrees of Freedom", 
    p_value  = "P-value"
    ) |> 
  tab_header(title = paste("Welch's ANOVA for log(Population) and Government Type")) |> 
  tab_style(style = cell_text(weight = "bold"), 
            locations = list(cells_title(groups = "title"), 
                             cells_column_labels()
                             )
            )
```

At an ⍺ = 0.05 significance level, we have extremely strong evidence that there is an association between the geometric mean population and the government types of countries.

Both Welch's ANOVA tests provide very strong evidence that *regime category* and *government type* are significantly associated with a country's population (p \< 0.001). Since large population countries may skew the mean, the median is a good indicator of typical country size across regime categories. Civilian dictatorships tend to have the greatest populations (median = 20,093,620), while parliamentary democracies have the smallest (median = 1,818,300). In terms of government type, communist countries have the largest median population by a large margin (25,950,000), with democratic monarchies being the smallest on average (2,116,427). Our ridgeline plots also reveal that distributions of country populations are more similar than distributions of log(GDP), with all distributions being fairly symmetrical. However, democratic monarchies and communist countries have bimodal features. While distributions are not identical across all political structures and equal variance was not satisfied in the model, *regime category* and *government type* are statistically significant predictors of a country's population, but should be interpreted with caution.

## An Analysis of GDP Across Political Ideologies

```{r}
#| fig-alt: "Ridgeline plots showing distributions of log(GDP) for different government types. 'Pure Monarchy' appears to have the highest median GDP, and 'Other' seems to have the lowest median GDP."

plot_gdp_by_category(gov_type)
```

The distributions of GDPs for pure monarchies, pure democracies, and other government types appear to be normal, while democratic monarchies and communist governments have a bimodal distribution. Pure monarchies or communist governments seem to have the largest average GDP, given the right-most center.

Communist countries show a staggeringly greater average GDP than other government types, at over \$3,000 billion, but the large standard error of 2,949 billion suggests that the difference may not be significant. Thus, looking at median GDP, pure monarchies are the greatest and other government types have the least GDP on average.

```{r}
summarize_gdp_by_category(regime_category)
```

The distributions of GDPs across regime categories appear to be normal, with the exception of parliamentary democracies, and mixed democratic regimes and military dictatorships are borderline. Royal dictatorships or presidential democracies appear to have the largest GDP on average, and military dictatorships appear to have the least.

Presidential democracies have the largest mean GDP, followed by parliamentary democracies and civilian dictatorships. The large standard errors suggests that these differences may not be significant, and median may be a better predictor of average GDP. The medians suggests that royal dictatorships have the largest average GDP.

```{r}
# Welch's ANOVA
gdp_regime_model <- oneway.test(log(GDP) ~ regime_category, data = country_meta_data, var.equal = FALSE)

data.frame(
    f_stat   = unname(gdp_regime_model$statistic),
    num_df   = unname(gdp_regime_model$parameter[1]),
    denom_df = unname(gdp_regime_model$parameter[2]),
    p_value  = gdp_regime_model$p.value
    ) |>
  gt() |> 
  cols_label(
    f_stat   = "F-statistic", 
    num_df   = "Numerator Degrees of Freedom", 
    denom_df = "Denominator Degrees of Freedom", 
    p_value  = "P-value"
    ) |> 
  tab_header(title = paste("Welch's ANOVA for log(GDP) and Regime Category")) |> 
  tab_style(style = cell_text(weight = "bold"), 
            locations = list(cells_title(groups = "title"), 
                             cells_column_labels()
                             )
            )

```

At an ⍺ = 0.05 significance level, we do not have enough evidence that there is an association between the geometric mean GDP and the regime categories of countries.

```{r}

# Welch's ANOVA
gdp_gov_model <- oneway.test(log(GDP) ~ gov_type, data = country_meta_data, var.equal = FALSE)

data.frame(
    f_stat   = unname(gdp_gov_model$statistic),
    num_df   = unname(gdp_gov_model$parameter[1]),
    denom_df = unname(gdp_gov_model$parameter[2]),
    p_value  = gdp_gov_model$p.value
    ) |>
  gt() |> 
  cols_label(
    f_stat   = "F-statistic", 
    num_df   = "Numerator Degrees of Freedom", 
    denom_df = "Denominator Degrees of Freedom", 
    p_value  = "P-value"
    ) |> 
  tab_header(title = paste("Welch's ANOVA for log(GDP) and Goverment Type")) |> 
  tab_style(style = cell_text(weight = "bold"), 
            locations = list(cells_title(groups = "title"), 
                             cells_column_labels()
                             )
            )

```

At an ⍺ = 0.05 significance level, we do not have enough evidence that there is an association between the geometric mean GDP and the government types of countries.

Both Welch's ANOVA tests do not provide enough evidence that *regime category* and *government type* are significantly associated with a country's GDP in US dollars in billions (p \> 0.1). However from our statistical summaries, we observed that presidential democracies have the largest nominal GDPs on average (mean = 776.25B), with military dictatorships having the lowest on average (mean = 73.25B). However, large GDP countries may skew the mean, so the median may be a better indicator of typical GDP. In this case, royal dictatorships have the greatest median nominal GDP (75.91B), with military dictatorships still being the lowest (11.48B). In terms of government type, pure monarchies have the largest median GDP (93.48B), with pure democracies being the lowest (34.36B), aside from *Other* government types. Our ridgeline plots also reveal that distributions of country log(GDP) vary widely from normally distributed, to skewed right (Communist), to bimodal (Democratic Monarchy). Since we cannot conclude that *regime category* nor *government type* are statistically significant predictors of a country's nominal GDP, we suggest that factors such as location, natural resources, and area may be greater correlated to GDP.
